---
title: "Gov 50 Final Project"
author: "Brit Shrader"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction 

Does incarceration act as an alternative to welfare or, alternatively, does increased welfare decrease incarceration? Scholars of mass incarceration and prison abolition argue that governments rely on prisons and policing to manage socioeconomic problems. Using this framework, I hypothesize that incarceration admission rates increase with decreased welfare spending, with both increased welfare spending decreasing poverty-associated criminalized acts and decreased welfare spending increasing state reliance on incarceration to address poverty. In this study, I specifically focus on the extent to which incarceration admission rates are associated with welfare spending per capita. Therefore, I expect that incarceration admission rates and welfare spending per capita will have a negative relationship. In other words, I expect that incarceration admission rates decrease with increased welfare spending per capita and incarceration admission rates increase with decreased welfare spending per capita. 

If I observe increased incarceration admission rates associated with decreased welfare spending per capita, this will support my hypothesis. However, if I observe decreased incarceration admission rates associated with decreased welfare spending per capita, this will provide evidence against my hypothesis. When I run my regression of the incarceration admission rates on the welfare spending per capita variable, a negative, significant coefficient would indicate support for my hypothesis. 

## Data 

```{r}
library(tidyverse)
library(infer)
library(broom)
library(dplyr)
library(modelsummary)
```

```{r}
## loading datasets

cspp_data_welfare <- read.csv("project_data/cspp_data_welfare.csv")

cspp_cleaned_welfare <- cspp_data_welfare |>
  select(st, year, apwspt)

cspp_data_pop <- read.csv("project_data/cspp_data_population.csv")

cspp_cleaned_pop <- cspp_data_pop |>
  select(st, year, poptotal, pop_annual)

load("project_data/NPPdata.rda")

npp_data <- da38249.0001

npp_cleaned_data <- npp_data |>
  select(STATE, YEAR,
         ADTOTM, ADTOTF)

```

```{r}
## renaming keys

npp_cleaned_data  <- npp_cleaned_data |>
  mutate(st = STATE, 
         year = as.integer(YEAR))

## merge data 

cspp_data <- cspp_cleaned_welfare |>
  left_join(cspp_cleaned_pop)

total_data <- cspp_data |>
  left_join(npp_cleaned_data)
```

```{r}
total_data <- total_data |> 
  mutate(admit = ADTOTM + ADTOTF,
         apwspt_cap = ifelse(is.na(poptotal) == FALSE, (apwspt/poptotal)*1000000, (apwspt/pop_annual)*1000000))
```

My sample is comprised of state level inmate admissions data and welfare spending per capita from 1977-2015. My units of analysis are state and year data. 

The explanatory variable of interest is welfare spending per capita. The variable is coded as a number representing the dollar amount spent on welfare per capita during a given year. For instance, if the variable were coded as 50, that would mean the state and local governments within the state spent 50 dollars per person on welfare total that year. For a given year, the amount the state and local governments spent on public welfare and veterans' services through direct expenditure is recorded in the Correlates of State Policy Project database <http://ippsr.msu.edu/public-policy/correlates-state-policy>. The welfare spending per capita was coded as this expenditure divided by the state population during the same year. This state population is also recorded in the Correlates of State Policy Project database. However, there are two variables recording this data. Thus, the variable used differs based on which is recorded in the given year.  

My outcome variable is the number of individuals admitted to a carceral facility in a state during a given year with a sentence of more than one year. For instance, if the variable were coded 50 that would mean 50 individuals with a sentence of more than one year were admitted to a carceral facility in the state during a given year. The number of male and female individuals with a sentence of more than one year were admitted to a local, federal, private, or state prison or jail within a state in a given year is recorded in the National Prisoner Statistics database <https://www.icpsr.umich.edu/web/NACJD/studies/38249>. The total admissions is coded as the sum of the male and female admissions. 

The data for this research is observational. While the data is recorded across a time period, my research design is cross-sectional because I do not compare changes over time. First, I perform a linear regression analysis with the incarceration admission rates as the dependent variable and welfare spending per capita as the independent variable. Due to possible time and state trends, I perform multiple linear regressions with the incarceration admission rates as the dependent variable and welfare spending per capita as an independent variable while controlling for year and state as additional independent variables. Additionally, I perform a linear multiple regression with incarceration admission rates as the dependent variable and state and year as the independent variables to examine the fit of the regression while excluding welfare spending per capita as an explanatory variable. 

## Data visualization 


```{r}
total_data |>
  ggplot(aes(x = year,
             y = ADTOTM + ADTOTF)) +
  geom_line(aes(group = st)) +
  geom_smooth(size = 1.5) +
  labs(x = "Year",
       y = "Total admissions",
       title = "Total inmate admissions per year by state, 1978-2015") +
  xlim(1978, 2015) +
  scale_y_continuous(labels = scales::comma)
```

## Results  

```{r}
lm_1 <- lm(admit ~ apwspt_cap, data = total_data)

lm_mult_1 <- lm(admit ~ apwspt_cap + factor(year) + factor(st), data = total_data)

lm_mult_2 <- lm(admit ~ apwspt_cap + factor(st), data = total_data)

lm_mult_3 <- lm(admit ~ apwspt_cap + factor(year), data = total_data)

lm_mult_4 <- lm(admit ~ factor(year) + factor(st), data = total_data)

var_labels <- c("(Intercept)" = "Intercept",
                "apwspt_cap" = "Welfare spending per cap")

modelsummary::modelsummary(list("Model 1 (welfare only)" = lm_1, 
                                "Model 2 (welfare, year, and state)" = lm_mult_1, 
                                "Model 3 (welfare and state)" = lm_mult_2, 
                                "Model 4 (welfare and year)" = lm_mult_3,
                                "Model 5 (state and year only)" = lm_mult_4),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           coef_map = var_labels,
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```

Model 1 conducts a regression on inmate admissions and welfare spending per capita without controlling for other variables. Model 1 suggests that an increase of one dollar in welfare spending per capita is associated with a increase of 3.84 inmate admissions, with a p-value of <0.001. However, the regression for Model 1 results in an adjusted R^2 value of 0.02. This adjusted R^2 value or estimated fit, which represents how much variation in inmate admissions can be explained by welfare spending per capita alone, is quite low. 

However, Model 2 conducts a multiple regression on inmate admissions and welfare spending per capita, controlling for state and year as confounding variables. It is important to control for state in this regression due to wide variation in inmate admission rates and welfare spending per capita by state. For example, California historically has both very high inmate admission rates and high welfare spending per capita compared to other states. Similarly, it is important to control for year in this regression inmate admission rates and welfare spending per capita both generally increase over time, but there are a number of other trends, such as inflation, that could explain this.

Moreover, Model 2 results in the highest adjusted R^2 out of all of the models with a value of 0.785. This high adjusted R^2 implies a better fit for the model. The welfare spending per capita coefficient is -6.09, with a standard error of 1.09. Thus, Model 2 suggests that, when holding state and year fixed, an increase of one dollar in welfare spending per capita is associated with a decrease of 6.09 (+/- 2.14) inmate admissions. With a p-value of <0.001 and a test level of $\alpha = 0.01$, we can reject the null hypothesis that the true coefficient for inmate admissions in this multiple regression is 0. 

Due to the nature of this study as both observational and cross-sectional, we cannot interpret the results of this regression casually. While I control for state and year, there may be other unobserved confounding variables. For example, I do not include any variables from the national level, such as national crime legislation or welfare spending. Additionally, I do not analyze how changes in welfare spending per capita influence incarceration admission rates over time. 


```{r}
total_data |>
  ggplot(aes(x = apwspt_cap,
             y = admit)) +
  geom_point(aes(color = year, alpha = .5)) +
  geom_smooth(method = "lm") +
   labs(x = "Welfare spending per cap",
       y = "Inmate admissions",
       title = "Inmate admission vs welfare spending, 1978-2019") +
  scale_y_continuous(labels = scales::comma)
```

## Conclusion

A conclusion section, including a recap of results, discussion of limitations of the study, and how to improve the study in the future (a paragraph)

In conclusion, when studying the relationship between incarceration admission rates and welfare spending per capita, the model that controls for state and year results in the best fit, based on its adjusted R^2. This model suggests that when holding state and year fixed, an increase of one dollar in welfare spending per capita is associated with a decrease of 6.09 (+/- 2.14) inmate admissions. While this provides support for my hypothesis that increased welfare spending decreases incarceration rates, I cannot make any casual claims about the relationship between incarceration admission rates and welfare spending per capita due to the limitations of this study. For this study, I must assume that these variables are the best representation of incarceration rates and welfare spending and that there is a linear relationship between these two variables. Additionally, I have to assume there are no unobserved confounding variables that influence this relationship. Due to the observational and cross-sectional nature of this study, it is not possible to make causal claims because I do not analyze how changes in welfare spending per capita influence incarceration admission rates. 

In order to improve this study in the future, it would be beneficial to adjust welfare spending per capita for inflation. It also may be beneficial to include incarceration rates for those admitted to a carceral facility with a sentence less than one year, since a large proportion of individuals are incarcerated in jails short-term. Additionally, to make casual claims, one could use a longitudinal before-and-after or difference-in-differences research design by conducting a linear regression where the independent variable is coded as the change in welfare spending from the previous year. Finally, one could conduct linear regression analyses with other variables that may be confounding, such as criminal sentencing laws, homelessness rates, and poverty rates.
